---
title: "EDA Project"
output:
  html_document:
    df_print: paged
---

This is the R Notebook for the project from Sifeng Xu, 24525844 for unit *CITS4009 Computation Data Analysis*. It demonstrates the process and findings of EDA based on the **Countries and Death Causes** dataset. 

```{r setup, echo=FALSE}
# Load  libraries
library(shiny)
library(shinyWidgets)
library(ggplot2)
library(gridExtra)
library(knitr)

# load dataset
df <- read.csv("./Countries and death causes.csv",header = T, sep=",")

```

## Basic information of the data


```{r glance, echo=TRUE}

data.frame(
  columeID = col(df)[1,],
  variables = names(df),
  class = sapply(df, typeof),
  first_values = sapply(df, function(x) paste0(head(x), collapse = ', ')),
  row.names = NULL) |>
kable()

summary(df) |> kable()

head(df) |> kable()

```


A first glance of data shows 31 columns which consists of 3 non-numerical data:

* countries' full name
* short country code
* year of measurements (1990-2019) 

the rest is numerical data showing how many deaths from each cause, each year and each country/region. Some countries are counted together such as *G20, OECD Countries, South-East Asia Region* and *Western Passific Region*. Also, some measured variables look similar such as *Outdoor.air.pollution* and *Air.pollution*.  

Most variables have zero values. This indicates that either no one died of that specific cause in a country in a specific year, or they were just missing values. Furthermore, there is no NA's in the dataset. 

```{r nas, echo=TRUE} 
print(sum(is.na(df)))

```

To give an impression what the dataset looks like, a snippet shows 6 countries and their deaths count from several causes in 2010.

```{r snippet, echo=TRUE}
df_2010 <- df[df['Year']==2010, 
  c('Entity', 'Code', 'Outdoor.air.pollution','Child.wasting', 'Smoking', 'Drug.use')]
countries <- c('Afghanistan', 'Morocco', 'Nigeria', 'Australia', 'Belgium', 'Congo')
snippet <- subset(df_2010, Entity %in% countries)
names(snippet)[1] <- 'Country'
snippet
```


## Plotting single variables


